Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-20 main.tex
also post deprecated
update

To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$ so that
\begin{align}
\label{eq:information}
H_{fa}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
\end{align}
By Gibbs inequality,
$$
H_{fa}(x, y;x', y', w') \le H_{fa}(x, y;x, y, y),
$$
though this bound can be made tighter.
In our simplifications we will make use of the constants
$$
a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
$$
and
$$
d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8
$$
with corresponding bounds
$$
0 \le a,b,c \le \frac{1}{2}
$$
and
$$
0 \le d \le \frac{1}{4}.
$$

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H_{fa}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H_{fa}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

>>>>

To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$ so that
\begin{align}
\label{eq:information}
H_{\tau'}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
\end{align}
By Gibbs inequality,
$$
H_{\tau'}(x, y;x', y', w') \le H_{\tau'}(x, y;x, y, y),
$$
though this bound can be made tighter.

\textbf{Show}: where these simplifications come about.
It's just adding/subtracting $\log$ terms, but could be good to walk through.

In our simplifications we will make use of the constants
$$
a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
$$
and
$$
d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8
$$
with corresponding bounds
$$
0 \le a,b,c \le \frac{1}{2}
$$
and
$$
0 \le d \le \frac{1}{4}.
$$

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H_{\tau'}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$
